Multilevel selection in causal models: the non-ergodicity of evolutionary and probability theories as the general driver of the emergence of cooperation and specialization.

### 2022-02-17 article.tex
published doc1.0.0

%opening
\title{\vspace{-2.2cm}Multilevel selection in causal models: the non-ergodicity of evolutionary and probability theories as the general driver for the emergence of cooperation and specialization.\footnote{\url{https://github.com/glandfried/transitions/releases/download/doc.0.0.0/landfried-trans\en{itions}\es{iciones}.pdf} } }
\author{Gustavo Landfried$^1$}
\affil{\small 1. Universidad de Buenos Aires. Facultad de Ciencias Exactas y Naturales. Departamento de Computaci\'on. Buenos Aires, Argentina}
\affil{\en{Mail}\es{Correo:} \url{glandfried@dc.uba.ar}}

@@

\begin{abstract}
\en{In the last third of the universe's history, a simple self-replicating organization of matter emerged on earth. }%
\es{En el último tercio de la historia del universo surgió en la tierra una organización de la materia simple capaz de autoreplicarse. }%
%
\en{The growth of these lineages followed multiplicative and noisy processes: sequences of survival and reproduction rates. }%
\es{El crecimiento de estos linajes siguieron procesos multiplicativos y ruidosos: secuencias de tasas de supervivencia y reproducción. }%
%
% \en{The errors produced during replication diversified the life forms, and the growth rates of the different strategies favored those better adapted to the environment. }%
% \es{Los errores producidos durante la replicación diversificaron las formas de vida, y las tasas de crecimiento de las diferentes estrategias favorecieron a aquellas mejor adaptadas al ambiente. }%
% %
\en{The current complexity of life is the consequence of a series of evolutionary transitions in which entities capable of self-replication after the transition become part of higher level cooperative units. }%
\es{La complejidad actual de la vida es consecuencia de una serie de transiciones evolutivas en las que entidades capaces de autoreplicación luego de la transición pasan a formar parte de unidades cooperativas de nivel superior. }%
%
% \en{How to explain this permanent tendency of life in favor of cooperative aggregation and specialization? }%
% \es{¿Cómo se explica esta tendencia permanente de la vida en favor de la agregación cooperativa y la especialización? }%
\en{To explain evolutionary transitions, it is necessary to demonstrate the evolutionary advantage of cooperation, but also the advantage of specialization. }%
\es{Para explicar las transiciones evolutivas es necesario demostrar la ventaja evolutiva de la cooperación, pero también la ventaja de la especiliazación. }%

@@

\en{Recently it has been shown that, as a consequence of the non-ergodicity of multiplicative processes, fluctuations have a negative effect on individual growth rates, which can be reduced through mutual cooperation, leading to an increase in the long-term growth rates of all individuals. }%
\es{Recientemente se ha mostrado que, como consecuencia de la no-ergodicidad de los procesos multiplicativos, las fluctuaciones tienen un efecto negativo en las tasas de crecimiento individuales, las cuales pueden ser reducidas a través de la mutua cooperación, produciendo un aumento de la tasas de crecimiento a largo plazo de todos los individuos. }%
%
% \en{Ole Peters believs that the increase in the growth rate is sufficient argument to demonstrate the evolutionary advantage of cooperation, which he proposes as the main explanation for evolutionary transitions. }%
% \es{Ole Peters considera que el aumento de la tasa de crecimiento es argumento suficiente para demostrar la ventaja evolutiva de la cooperación, lo que propone como principal explicación de las transiciones evolutivas. }%
% %
% \en{However, he does not consider the problem of defection, who says ``our cooperators are unable to break the cooperative pact''. }%
% \es{Sin embargo no considera el problema de la deserción, quien dice ``our cooperators are unable to break the cooperative pact''. }%
% %
% \en{Yaari suggests a group selection argument but he did not provide a formal demonstration. }
% \es{Yaari sugiere un argumento de selección de grupos pero no prové una demostración formal. }
\en{However, these papers do not provide a formal proof of the evolutionary advantage of cooperation. }%
\es{Sin embargo estos trabajos no porveen una demostración formal que justifique la ventaja evolutiva de la cooperación. }%
% Parrafo
%
\en{For this purpose it is necessary to consider selection at both the individual and group level. }%
\es{Para este propósito es necesario considerar selección tanto a nivel individual como a nivel grupal. }%
%
\en{The co-author of the concept of evolutionary transitions (Szathmáry) recently proposed to analyze the evolution of populations subject to multilevel selection by means of Bayesian hierarchical models, making use of the isomorphism between evolutionary theory and Bayesian inference. }%
\es{El co-autor del concepto de transiciones evolutivas (Szathmary) propuso recientemente analizar la evolución de las poblaciones sujetas a selección multinivel mediante modelos jerárquicos bayesianos, haciendo uso del isomorfismo entre las teoría de la evolución y la inferencia bayesiana. }%
%
\en{However, this work does not provide any model that performs multilevel selection, so the proposal remains open. }%
\es{Sin embargo, este trabajo no proveé ningun modelo que realice selección multinivel, por lo que la propuesta sigue abierta. }%

@@

\en{In this paper we specify a probabilistic causal model, in which individuals are affected by the environment and by the social behaviors of cooperation and defection of their context. }%
\es{En este trabajo especificamos un modelo causal probabilístico, en el que los individuos se ven afectados por el ambiente y por los comportamientos sociales de cooperación y deserción de su contexto. }%
\en{Under this minimal set of hypotheses, where we consider \emph{unconditionally} cooperative individuals who generate a common good that can be exploited by defecting individuals without receiving some kind of punishment in return (e.g. end of cooperation), the evolution of cooperation literature predicts defection as the only evolutionarily stable strategy. }%
\es{Bajo este conjunto mínimo de hipótesis, donde consideramos individuos \emph{incondicionalmente} cooperadores que generan un bien común que puede ser explotado por individuos desertores sin que reciban a cambio algún tipo de castigo (e.g fin de la cooperación), la literatura de evolución de la cooperación predice que la deserción es la única estrategia evolutivamente estable. }%
% \en{It is assumed that common goods, if not accompanied by special conditions (such as communication allowing coordination, memory allowing rewards or punishments, etc.), lead to their overexploitation, because even if the optimum is obtained through mutual cooperation there would be an individual incentive to defect. }%
% \es{Se supone que los bienes comunes, si no están acompañados de condiciones especiales (como la comunicación que permita la coordinación, la memoria que permita aplicar premios o castigos, etc), conducen a su sobreexplotación, porque aunque el óptimo se obtenga a través de la mutua cooperación habría una incentivo individual para desertar. }%
\en{However, probabilistic inference shows that cooperative individuals are favored by multilevel selection. }%
\es{Sin embargo, la inferencia probabilística muestra que los individuos cooperadoras se ven favorecidas mediante selección multinivel. }%
\en{While resource-avoidance strategies can invade entirely cooperative groups by natural selection, such behavior at the same time increases their own individual fluctuations, reducing their own long-term growth rate without the need to introduce penalties. }%
\es{Si bien las estrategias que evitan compartir recursos pueden invadir por selección natural grupos enteramente cooperadores, tal comportamiento aumenta sus propias fluctuaciones individuales, reduciendo su propia tasa de crecimiento a largo plazo sin necesidad de introducir castigos. }%
\en{That is, contrary to the belief established since the mid-20th century in economics, we show that the dynamics of common goods cannot be represented by a prisoner's dilemma payoff matrix. }%
\es{Es decir, en contra de la creencia establecida que se tiene en economía desde mediados del siglo 20, mostramos que los dinámicas de bienes comunes no puede representarse mediante una matriz de pagos del dilema del prisionero. }%
\en{Moreover, contrary to the belief that specialization is too complex a feature to produce a benefit in simple aggregations, we show that as soon as cooperation emerges, an advantage in favor of specialist strategies appears even in groups of size 2. }%
\es{Además, en contra de la creencia de que la especialización es una caracterísitica demasiado compleja para que produzca un beneficio en agregaciones simples, mostramos que apenas surge la cooperación, aparece una ventaja a favor de las estrategias especialistas incluso en grupos de tamaño 2. }%
\en{Since the specialist strategies are individually poorly adapted to the environment, an irreversibility of the evolutionary transition is created. }%
\es{Como la estrategias especialistas están individualmente mal adaptadas al ambiente, se crea una irreversbilidad de las transiciones evolutivas. }%
\en{Finally, although in this causal model there is no evolutionary advantage in favor of the division of labor, its emergence is viable by evolutionary drift. }%
\es{Finalmene, si bien en este modelo causal no hay una ventaja evolutiva a favor de la división del trabajo, su aparición es viable por deriva evolutiva. }%

\en{The reason why an advantage in favor of cooperation and specialization arises in our simple causal model is due to the multiplicative (non-ergodic) nature of probability theory and its isomorphism with evolutionary theory. }%
\es{El motivo por el cual surge una ventaja a favor de la cooperación y la especialización en simple modelo causal se debe a la naturaleza multiplicativa (no-ergódica) de la teoría de la probabililidad y a su isomorfismo con la teoría de la evolución. }%
%
% \en{Even if our causal model does not assume any kind of process to update resources, probabilistic inference is proportional to the resources obtained through a multiplicative process. }%
% \es{Incluso si nuestro modelo causal no presupone ningún tipo de proceso para actualizar los recursos, la inferencia probabilistica es proporcional a los recursos obtenidos a través de un proceso multiplicativo. }%
% %
\en{The multiplicative updating of the probabilities of individuals, which arises naturally from applying the rules of probability to the causal model, is in line with the long-established idea in evolutionary theory that the growth of lineages follow multiplicative processes. }%
\es{La actualización multiplicativa de las probabilidades de los individuos, que surge naturalmente de aplicar las reglas de la probabilidad al modelo causal, está en línea con la idea largamente establecida en la teoría de la evolución de que el crecimiento de los linajes siguen procesos multiplicativos. }% 
%

@@

\en{In this section we present the isomorphism between probabilitic and evolutionary theory. }%
\es{En esta sección presentamos el isomorfismo entre las teorías de la probabilidad y de la evolución. }%
%
\en{In the \ref{sec:resuts} section we will poof the evolutionary advantage of cooperation and specialization using a Bayesian hierarchical model representing evolution under multilevel selection. }%
\es{En la sección \ref{sec:results} demostraremos la ventaja evolutiva de la cooperación en presencia de deserción mediante un modelo jerárquico bayesiano que representa evolución bajo selección multinivel. }%
%

@@

\en{Probability theory can be summarized in two rules: the \ref{eq:sum_rule} and the~\ref{eq:product_rule}. }%
\es{La teoría de la probabilidad puede resumirse en dos reglas: la~\ref{eq:sum_rule} y la~\ref{eq:product_rule}. }%
%
\en{The \ref{eq:sum_rule} states that any marginal distribution can be obtained by integrating or summing the joint distribution. }%
\es{La \ref{eq:sum_rule} afirma que cualquier distribución marginal se puede obtener integrando o sumando la distribución conjunta. }%
%
\begin{equation} \label{eq:sum_rule}
 \tag{\en{sum rule}\es{regla de la suma}}
 P(x) = \sum_{y} P(x,y) \ \ \ \ \ \text{or} \ \ \ \ \ p(x) = \int p(x,y) \, dy
\end{equation}
%
\en{Where usually $P(\cdot)$ and $p(\cdot)$ represent discrete and continuous probability distributions respectively. }%
\es{Donde usualmente $P(\cdot)$ y $p(\cdot)$ representan distribuciones de probabilidad discretas y continuas respectivamente. }%
%
\en{On the other hand, the \ref{eq:product_rule} states that any joint distribution can be expressed as the product of one-dimensional conditional distributions. }%
\es{Por su parte, la \ref{eq:product_rule} se\~nala que cualquier distribuci\'on conjunta puede ser expresada como el producto de distribuciones condicionales uni-dimensionles. }%
%
\begin{equation}\label{eq:product_rule}
\tag{\en{product rule}\es{regla del producto}}
 P(x,y,z) = P(x|y,z) P(y|z) P(z)
\end{equation}
%
\en{This expression can be simplified by eliminating from the conditionals the variables that do not affect the function. }%
\es{Esta expresión se puede simplificar eliminado de los condicionales las variables que no afectan a la función. }%
%
\en{The minimum set of variables that appear on the right side of the conditional are said to be the ``parents'' of the left side variable. }%
\es{Al conjunto mínimo de variables que aparecen del lado derecho del condicional decimos que son los ``padres'' de la variable de la izquierda. }%

\en{Another way of defining joint distributions is by means of graphical models, in which there is a node for each variable, and an arrow for each pair ``parent$\rightarrow$child''. }%
\es{Otra forma de definir distribuciones conjuntas es mediante modelos gráficos, en los que hay un nodo por cada variable, y una flecha por cada par ``padre$\rightarrow$hijo''. }% 
%
\en{Graphical models are a well-established tool in probability because they: 1. provide an intuitive language to unambiguously express all hypotheses, 2. reduce the dimensionality of the joint probability distribution, 3. and allows to perform inference efficiently through the sum-product algorithm~\cite{kschischang2001-factorGraphsAndTheSumProductAlgorithm}. }%
\es{Los modelos gráficos son una herramienta bien establecida en probabilidad debido a que: 1. ofrecen un lenguaje intuitivo para expresar sin ambigüedad todas las hipótesis; 2. reducen la dimensionalidad de la distribución de probabilidad conjunta; 3. y permiten realizar inferencia de forma eficiente a través del algoritmos suma-producto~\cite{kschischang2001-factorGraphsAndTheSumProductAlgorithm}. }%
\en{To use the graphical models no causal interpretation is required, in fact the conditional probabilities can be expressed in any order. }%
\es{Para utilizar los modelos gráficos no se requiere una interpretación causal, de hecho las probabilidades condicionales pueden expresar en cualquier orden. }%

@@

\label{fig:czegel_et_el}
\end{figure}
%
\en{The validity of this model is trivial as the product rule always allows the following decomposition, }%
\es{La validez de este modelo es trivial en tanto la regla del producto siempre permite la siguente descomposición, }%
%
\begin{equation}
P(\Aa,i,c^1,c^2)= P(\Aa|i,c^1,c^2)P(i|c^1,c^2)P(c^1|c^2)P(c^2)
\end{equation}
%
\en{However, the model does not express the causal relationship between the variables. }%
\es{Sin embargo el modelo no expresa la relación causal entre las variables. }%
%
\en{This is evident in all conditional probability distributions. }%
\es{Esto se constata en todas las distribuciones de probabilidad condicional. }%
%
\en{At one extreme, the level 2 collective $C^2$, which is composed of both the level 1 collective and individuals, appears in the model as an independent variable. }%
\es{En un extremo, el colectivo de nivel 2 $C^2$, que está compuesta tanto por el colectivo de nivel 1 y por los individuos, aparece en el modelo como una variable independiente. }%
%
\en{At the other extreme, the environment $\A$, which is usually considered in evolutionary causal models as an independent random variable, appears in this model as dependent on individuals and collectives. }%
\es{En el otro extremo, el ambiente $\A$, que suele considerarse en los modelos causales evolutivos como una variable aleatoria independiente, aparece en este modelo como dependiente de los individuos y los colectivos. }%
%
\en{Conversely, all collectives and individuals appear independent of the environment. }%
\es{Y a la inversa, todas los colectivos e individuos aparecen independientes del ambiente. }%
%
\en{This perhaps explains why the authors avoided specifying conditional probability distributions, and instead provided a pictorial example of the joint probability distribution. }%
\es{Esto quizás explique porque los autores evitaron especificar las distribuciones de probabilidad condicional, y ofrecieron en cambio un ejemplo pictórico de la distribución de probabilidad conjunta. }%

\subsection{\en{Basic causal model}\es{Modelo causal básico}}

\en{The basic causal model we present here serves to introduce the methodology we will use in section \ref{sec:results}, and the conclusion we will draw from it will be of interest later on. }%
\es{El modelo causal básico que presentamos aquí sirve para introducir la metodología que utilizaremos en la sección \ref{sec:results}, y la conclusión que de él saquemos será de interés más adelante. }%
% \en{To define conditional probability distributions it is convenient to express the models in causal terms. }%
% \es{Para definir las distribuciones de probabilidad condicional es conveniente expresar los modelos en términos causales. }%
\en{This model generalizes the problems analyzed by Lewontin, Yaari and Peters~\cite{lewontin1969-randomlyVaryingEnvironment, peters2019-ergodicityEconomi.s}, and formalizes in probabilistic terms the model proposed by Yaari-Solomon~\cite{yaari2010-cooperationEvolution}. }%
\es{Este modelo generaliza los problemas analizados por Lewontin-Cohen y Peters~\cite{lewontin1969-randomlyVaryingEnvironment, peters2019-ergodicityEconomics}, y formaliza en términos probabilisticos el modelo propuesto por Yaari-Solomon~\cite{yaari2010-cooperationEvolution}. }% 
\en{In them, the binary environmental states depend on a probability $p$. }%
\es{En ellos, los estados ambientales binarios dependen de un probabilidad $p$. }%

@@

P(\Aa) = p^{\Aa} (1-p)^{(1-\Aa)}

@@

\en{To define the probability of the strategies, we need a space of alternative strategies, not present in the Lewontin-Cohen and Peters-Amadou models. }%
\es{Para definir la probabilidad de las estrategia, necesitamos un espacio de estrategias alternativas, no presente en los modelo de Lewontin-Cohen y de Peters-Amadou. }%

@@

\en{While we are free to propose any set of strategies, we will consider only those that allocate a limited resource $R$ among the binary states of the environment, with $ 0 \leq \Ee \leq R$. }%
\es{Si bien tenemos la libertad de proponer cualquier conjunto de estrategias, vamos a considerar sólo a aquellas que reparten un recurso limitado $R$ entre los estados binarios del ambiente, con $ 0 \leq \Ee \leq R$. }%

@@

\en{The strategy proposed by Lewontin-Cohen is $\Ee=1.7$ with $R=2.2$, and the strategy proposed by Peters is $\Ee=1.5$ with $R=2.1$. }%
\es{La estrategia propuesta por Lewontin-Cohen es $\Ee=1.7$ con $R=2.2$, y la de Peters es $\Ee = 1.5$ con $R=2.1$. }%

@@

\en{To express all this strategies with a single parameter, we will work only with the family of fitnesses in which $R=1$, $f(\Ee, \Aa)$. }%
\es{Para expresar las estrategias con un sólo parámetro, vamos a trabajar solamente con la familia de aptitudes en la que $R=1$, $f(\Ee, \Aa)$. }%

@@

\begin{equation}\label{eq:familia_de_aptitudes}
f(\Ee, \Aa) = \begin{cases}
 \Ee & \Aa = 1 \\
 1-\Ee & \Aa = 0

>>>>

%opening
\title{\vspace{-2.2cm}Multilevel selection in causal models: the multiplicative nature of evolutionary and probabilistic selection processes as the general driver for the irreversibility emergence of cooperation and specialization.\footnote{\url{https://github.com/glandfried/transitions/releases/download/doc1.0.0/landfried-transitions.pdf}} }
\author{Gustavo Landfried$^1$}
\affil{\small 1. Universidad de Buenos Aires. Facultad de Ciencias Exactas y Naturales. Departamento de Computaci\'on. Buenos Aires, Argentina}
\affil{\en{Mail}\es{Correo:} \url{glandfried@dc.uba.ar}}

@@

\begin{abstract}

\en{To explain why major evolutionary transitions are so common in the history of life we need to find the causes that systematically generate the irreversible emergence of cooperation and specialization. }%
\es{Para explicar por qué las grandes transiciones evolutivas son tan comunes en la historia de la vida necesitamos encontrar las causas que sistemáticamente generan la emergencia irreversible de la cooperación y la especialización. }%
%
\en{For this purpose it is necessary to consider selection at both the individual and group level. }%
\es{Para este propósito es necesario considerar selección tanto a nivel individual como a nivel grupal. }%
%
\en{The co-author of the concept of major evolutionary transitions (Szathmáry) recently proposed to analyze the evolution of populations subject to multilevel selection by means of Bayesian hierarchical models, making use of the isomorphism between evolutionary theory and Bayesian inference. }%
\es{El co-autor del concepto de transiciones evolutivas mayores (Szathmáry) propuso recientemente analizar la evolución de las poblaciones sujetas a selección multinivel mediante modelos jerárquicos bayesianos, haciendo uso del isomorfismo entre las teoría de la evolución y la inferencia bayesiana. }%
%
\en{However, the proposal remains open. }%
\es{Sin embargo, la propuesta sigue abierta. }%
%
% \en{Moreover, the models do not express a causal relationship between the variables. }%
% \es{Además, los modelos no expresan una relación causal entre las variables. }%

@@

\en{In this paper we specify a probabilistic causal model, in which individuals are affected by the environment and by the social behaviors of cooperation and defection of their context. }%
\es{En este trabajo especificamos un modelo causal probabilístico, en el que los individuos se ven afectados por el ambiente y por los comportamientos sociales de cooperación y deserción de su contexto. }%
%
\en{Even under this minimal set of hypotheses, where we consider \emph{unconditionally} cooperative individuals who generate a common good that can be exploited by defecting individuals without receiving some kind of punishment in return (e.g. end of cooperation), probabilistic inference shows that cooperative individuals are favored by multilevel selection. }%
\es{Incluso bajo este conjunto mínimo de hipótesis, donde consideramos individuos \emph{incondicionalmente} cooperadores que generan un bien común que puede ser explotado por individuos desertores sin que reciban a cambio algún tipo de castigo (e.g fin de la cooperación), la inferencia probabilística muestra que los individuos cooperadoras se ven favorecidas mediante selección multinivel. }%
%
\en{In addition, we show that as soon as cooperation emerges, an advantage in favor of specialist strategies appears. }%
\es{Además, mostramos que apenas surge la cooperación, aparece una ventaja a favor de las estrategias especialistas. }%
%
\en{Since the specialist strategies are individually poorly adapted to the environment, an irreversibility of the evolutionary transition is created. }%
\es{Como la estrategias especialistas están individualmente mal adaptadas al ambiente, se crea una irreversbilidad de la transici evolutivas. }%

@@

\en{The reason why an advantage in favor of cooperation and specialization arises in our simple causal model is due to the multiplicative (non-ergodic) nature of probability theory and its isomorphism with evolutionary theory. }%
\es{El motivo por el cual surge una ventaja a favor de la cooperación y la especialización en simple modelo causal se debe a la naturaleza multiplicativa (no-ergódica) de la teoría de la probabililidad y a su isomorfismo con la teoría de la evolución. }%
% %
% \en{The multiplicative updating of the probabilities of individuals, which arises naturally from applying the rules of probability to the causal model, is in line with the long-established idea in evolutionary theory that the growth of lineages follow multiplicative processes. }%
% \es{La actualización multiplicativa de las probabilidades de los individuos, que surge naturalmente de aplicar las reglas de la probabilidad al modelo causal, está en línea con la idea largamente establecida en la teoría de la evolución de que el crecimiento de los linajes siguen procesos multiplicativos. }%

@@

\en{In this section we present the isomorphism between probabilitic and evolutionary theory. }%
\es{En esta sección presentamos el isomorfismo entre las teorías de la probabilidad y de la evolución. }%
%
\en{In section ``\nameref{sec:results}'' we will prove the evolutionary advantage of cooperation, specialization and their irreversibility using a Bayesian hierarchical model with a causal interpretation representing evolution under multilevel selection. }%
\es{En la sección ``\nameref{sec:results}'' demostraremos la ventaja evolutiva de la cooperación, la especialización y su irreversbilidad mediante un modelo jerárquico bayesiano con una interpretación causal que representa evolución bajo selección multinivel. }%
%

@@

\en{Probability theory can be summarized in two rules: the sum rule and the product rule. }%
\es{La teoría de la probabilidad puede resumirse en dos reglas: la regal de la suma y la regla del producto. }%
%
\en{The product rule allows describing multidimensional models through causal mechanisms: belief distributions on each effect, given the hidden values of its causes (conditional probabilities). }%
\es{La regla del producto permite describir modelos multidimensionales a través de mecanismos causales: distribuciones de creencias sobre cada uno de los efectos, dados los valores ocultos de sus causas (probabilidades condicionales). }%
%
\en{The sum rule, on the other hand, predicts the behavior of a variable (marginal probabilities) with the contribution of all the hypotheses of the multidimensional causal model. }%
\es{La regla de la suma, por su parte, predice el comportamiento de una variable (probabilidades marginales) con la contribución de todas las hipótesis contradictorias del modelo causal multidimensional. }%
% Parrafo
\en{Graphical models are a well-established tool in probability because they: 1. provide an intuitive language to unambiguously express all hypotheses, 2. reduce the dimensionality of the joint probability distribution, 3. and allows to perform inference efficiently through the sum-product algorithm~\cite{kschischang2001-factorGraphsAndTheSumProductAlgorithm}. }%
\es{Los modelos gráficos son una herramienta bien establecida en probabilidad debido a que: 1. ofrecen un lenguaje intuitivo para expresar sin ambigüedad todas las hipótesis; 2. reducen la dimensionalidad de la distribución de probabilidad conjunta; 3. y permiten realizar inferencia de forma eficiente a través del algoritmos suma-producto~\cite{kschischang2001-factorGraphsAndTheSumProductAlgorithm}. }%
\en{To use the graphical models no causal interpretation is required, in fact the conditional probabilities can be expressed in any order. }%
\es{Para utilizar los modelos gráficos no se requiere una interpretación causal, de hecho las probabilidades condicionales pueden expresar en cualquier orden. }%

@@

\label{fig:czegel_et_el}
\end{figure}
%
\en{The validity of this model is trivial as the product rule always allows the following decomposition, }%
\es{La validez de este modelo es trivial en tanto la regla del producto siempre permite la siguente descomposición, }%
%
\begin{equation}
P(\Aa,i,c^1,c^2)= P(\Aa|i,c^1,c^2)P(i|c^1,c^2)P(c^1|c^2)P(c^2)
\end{equation}
%
\en{However, the model does not express the causal relationship between the variables. }%
\es{Sin embargo el modelo no expresa la relación causal entre las variables. }%
%
\en{This is evident in all conditional probability distributions. }%
\es{Esto se constata en todas las distribuciones de probabilidad condicional. }%
%
\en{At one extreme, the level 2 collective $C^2$, which is composed of both the level 1 collective and individuals, appears in the model as an independent variable. }%
\es{En un extremo, el colectivo de nivel 2 $C^2$, que está compuesta tanto por el colectivo de nivel 1 y por los individuos, aparece en el modelo como una variable independiente. }%
%
\en{At the other extreme, the environment $\A$, which is usually considered in evolutionary causal models as an independent random variable, appears in this model as dependent on individuals and collectives. }%
\es{En el otro extremo, el ambiente $\A$, que suele considerarse en los modelos causales evolutivos como una variable aleatoria independiente, aparece en este modelo como dependiente de los individuos y los colectivos. }%
%
\en{Conversely, all collectives and individuals appear independent of the environment. }%
\es{Y a la inversa, todas los colectivos e individuos aparecen independientes del ambiente. }%
%
% \en{This perhaps explains why the authors avoided specifying conditional probability distributions, and instead provided a pictorial example of the joint probability distribution. }%
% \es{Esto quizás explique porque los autores evitaron especificar las distribuciones de probabilidad condicional, y ofrecieron en cambio un ejemplo pictórico de la distribución de probabilidad conjunta. }%

\subsection{\en{A basic causal model: advantage for generalist individuals}\es{Un modelo causal básico: ventaja para los individuos generalistas}}

\en{The basic causal model we present here serves to introduce the methodology we will use in section ``\nameref{sec:results}'', and the conclusion we will draw from it will be of interest later on. }%
\es{El modelo causal básico que presentamos aquí sirve para introducir la metodología que utilizaremos en la sección ``\nameref{sec:results}'', y la conclusión que de él saquemos será de interés más adelante. }%
%
\en{In them, the binary environmental states depend on a probability $p$ (for example $p$ is the probability of the coin coming up heads). }%
\es{En ellos, los estados ambientales binarios dependen de un probabilidad $p$ (por ejemplo $p$ es la probabilidad de la moneda de salir cara). }%
%
\begin{equation}
P(\Aa) = p^{\Aa} (1-p)^{(1-\Aa)}
\end{equation}
%
\en{While we are free to propose any set of strategies, we will consider only those that allocate a limited resource $R$ among the binary states of the environment, with $ 0 \leq \Ee \leq R$. }%
\es{Si bien tenemos la libertad de proponer cualquier conjunto de estrategias, vamos a considerar sólo aquellas que reparten un recurso limitado $R$ entre los estados binarios del ambiente, con $ 0 \leq \Ee \leq R$. }%
%
\en{The strategy proposed by Lewontin-Cohen is $\Ee=1.7$ with $R=2.2$, and the strategy proposed by Peters is $\Ee=1.5$ with $R=2.1$. }%
\es{La estrategia propuesta por Lewontin-Cohen es $\Ee=1.7$ con $R=2.2$, y la de Peters es $\Ee = 1.5$ con $R=2.1$. }%
%
\en{To express all this strategies with a single parameter, we will work only with the family of fitnesses in which $R=1$, $f(\Ee, \Aa)$. }%
\es{Para expresar las estrategias con un sólo parámetro, vamos a trabajar solamente con la familia de aptitudes en la que $R=1$, $f(\Ee, \Aa)$. }%
%
\begin{equation}\label{eq:familia_de_aptitudes}
f(\Ee, \Aa) = \begin{cases}
 \Ee & \Aa = 1 \\
 1-\Ee & \Aa = 0
  \end{cases}
\end{equation}
%
\en{Now, all strategies lie between 0 and 1: the strategy $\Ee = 1.5/2.1 \approx 0.71$ is the one proposed by Peters, and the strategy $\Ee = 1.7/2.2 \approx 0.77$ is the one proposed by Lewontin-Cohen. }%
\es{Ahora, todas las estrategias están entre 0 y 1: la estrategia $\Ee = 1.5/2.1 \approx 0.71$ es la propuesta por Peters, y la estrategia $\Ee = 1.7/2.2 \approx 0.77$ es la propuesta por Lewontin-Cohen. }%
%
\en{For other ways of defining this same function, see Yaari-Solomon equation 11~\cite{yaari2010-cooperationEvolution}. }%
\es{Para ver otras formas de definir esta misma función, ver la ecuación 11 de Yaari-Solomon~\cite{yaari2010-cooperationEvolution}. }%
%
\en{To define the probability of the strategies given the environment, $P(\Ee|\Aa)$, we need the set of strategies to integrate 1. }%
\es{Para definir la probabilidad de las estrategias dado el ambiente, $P(\Ee|\Aa)$, necesitamos que el conjunto de estrategias integren 1. }%
%
\en{Let's say, for now, that the probability of the strategy is proportional to its fitness, }%
\es{Digamos, por ahora, que la probabilidad de la estrategia es proporcional a su aptitud, }%
%
\begin{equation}\label{eq:probabilidad_propro_aptitud}
P(\Ee | \Aa) \propto  \Ee^{\Aa}(1-\Ee)^{1-\Aa} = f(\Ee,\Aa)
\end{equation}
%
\en{This allows us to partially define a graphical model with a causal interpretation, }%
\es{Esto nos permite definir parcialmente un modelo gráfico con una interpretación causal, }%
%
\begin{figure}[ht!]
\centering
\tikz{
    \node[latent] (e) {$\A_t$};
    \node[const, right=of e] (en) {\ $P(\Aa) = p^{\Aa} (1-p)^{(1-\Aa)}$};
    \node[const, left=of e] (ne) {\en{Environment}\es{Ambiente}: \ \ \ };

    \node[latent, below=of e] (r) {$\E$};
    \node[const, right=of r] (rn) {$P(\Ee|\Aa) \propto f(\Ee,\Aa)$};
    \node[const, left=of r] (nr) {\en{Strategy}\es{Estrategia}: \ \ \ };

    \edge {e} {r};
    \plate {ee} {(e)} {$t$}; 
    }
\caption{
 \en{Causal model in which the environment affects individuals. }%
 \es{Modelo causal en el que el ambiente afecta a los individuos. }%
 }
\label{fig:model-lewontin-peters}
\end{figure}
%
\en{in which the environment is an independent random variable and the strategies depends on the observed states of the environment. }%
\es{en el que el ambiente es una variable aleatoria independiete y las estrategias dependen de los estados observados del ambiente. }%
%
\en{The result of the posterior will indicate the evolutionary stability of the strategies. }%
\es{El resultado del posterior nos indicará la estabilidad evolutiva de las estrategias. }%

\en{Suppose that the environmental states are generated with probabilities $P(\A = 1) = 0.71$ and $P(\A = 0) = 0.29$. }%
\es{Supongamos que los estados del ambiente se generan con probabilidades $P(\A = 1) = 0.71$ y $P(\A = 0) = 0.29$. }%
%
\en{In Figure~\ref{fig:estrategias_individuales} we show how the posterior of the strategies changes as we add observations to the model. }%
\es{En la figura~\ref{fig:estrategias_individuales} mostramos cómo cambia el posterior de las estrategias a medida que agregamos observaciones al modelo. }%
%
\begin{figure}[H]
    \centering
    \begin{subfigure}[b]{0.32\textwidth}
    \includegraphics[width=\linewidth]{figures/coin1.pdf}
    \caption{$T = 1$}
    \end{subfigure}
    \begin{subfigure}[b]{0.32\textwidth}
    \includegraphics[width=\linewidth]{figures/coin2.pdf}
    \caption{$T = 10$}
    \end{subfigure}
    \begin{subfigure}[b]{0.32\textwidth}
    \includegraphics[width=\linewidth]{figures/coin3.pdf}
    \caption{$T = 10^5$}
    \end{subfigure}
    \caption{
    \en{Posterior probability of strategies as time progresses ($T=1, \, T=10, \, T=10^5$). }%
    \es{Probabilidad posterior de las estrategias a medida que avanza el tiempo ($T=1, \, T=10, \, T=10^5$). }%
    }
    \label{fig:estrategias_individuales}
\end{figure}
%
\en{The evolutionary process selects the most generalist individual strategy of all, the one that distributes resources in the same proportion as the states of the environment are generated. }%
\es{El proceso evolutivo selecciona la estrategia individual más generalista de todas, la que reparte los recursos en la misma proporción que se generan los estados del ambiente. }%
%
\en{When the environment generates the states with probability $p=0.71$, then the best adapted strategy is $\Ee = 0.71$. }%
\es{Cuando el ambientes genera los estados con una probabilidad $p=0.71$, entonces la estrategia mejor adaptada es $\Ee = 0.71$. }%
%
\en{The selection of individual strategies, now intuitive, will be different and perhaps counter-intuitive when we incorporate the possibility of cooperation and defection into the model. }%
\es{La selección de estrategias individuales, ahora intuitiva, será diferente y quizás contra-intuitiva cuando incorporemos al modelo la posibilidad de cooperación y deserción. }%

% \en{In this section we will analyze whether there is indeed any evolutionary advantage of cooperation and specialization in the presence of defection. }%
% \es{En esta sección analizaremos si efectivamente existe alguna ventaja evolutiva de la cooperación y la especialización en presencia de deserción. }%
% %
\en{In this section we will extend the causal model proposed in section ``\nameref{sec:methodology}'' by incorporating unconditionally cooperative behaviors, such as the one presented in the ``\nameref{sec:introduction}'' section, including defecting behaviors (which receive the benefit of cooperation without contributing to it). }%
\es{En esta sección extenderemos el modelo causal propuesto en la sección ``\nameref{sec:methodology}'' incorporando comportamientos incondicionalmente cooperativos, como el presentado en la sección ``\nameref{sec:introduction}'', incluyendo comportamientos desertores (que reciben el beneficio de la cooperación sin aportar al mismo). }%
\en{and if their social behavior is cooperative, $\texttt{coop}(i) =  i \ \texttt{mod} \ N < \texttt{region}(i)$ (the first $n$ individuals in the region are cooperators and the rest defectors); }%
\es{y si su comportamiento social es cooperador, $\texttt{coop}(i) =  i \ \texttt{mod} \ N < \texttt{region}(i)$ (los primeros $n$ individuos de la región son cooperadores y el resto desertores); }%

@@

@@

@@

\en{As we have seen in the section ``\nameref{sec:methodology}'', the probability of the environmental states is, }%
\es{Como hemos visto en la sección ``\nameref{sec:methodology}'', la probabilidad de los estados ambientales es, }%

@@

@@

\en{but now at each time $t$ we have a vector $\vec{\Aa}$, in which the $i-$th environmental state influences the $i-$th individual, }%
\es{pero ahora en cada tiempo $t$ tenemos un vector $\vec{\Aa}$, en el que el $i-$ésimo estado ambiental influencia al $i-$ésimo individuo, }%

@@
